\documentclass{article}
\usepackage[margin=1.25in]{geometry}
\usepackage{fancyhdr}
\usepackage{listings}
\pagestyle{fancy}
\lhead{} \rhead{}
\chead{Splice Machine Technical Design Document}
\lfoot{Splice Machine, Inc. Proprietary and Confidential}
\cfoot{}
\rfoot{\thepage\\}

\begin{document}

\title{Authentication and Authorization}
\author{John Leach}
\maketitle
\makeauthor
\thispagestyle{fancy}

\section{Revision History}
\begin{enumerate}
  \item 7/3/2014 - (JL) Initial version
  \item 7/25/2014 - (GD) Minor updates to impact, etc.
  \item 8/22/2014 - (CZ) Update some wording, LDAP, add config settings, etc.
\end{enumerate}

\section{Background}
Transactional and Analytical databases are required to provide an
authentication and authorization mechanism.

\section{Requirements}

\subsection{P0 Requirements}
\begin{itemize}
  \item Basic Authentication/Authorization as provided by Apache Derby's 10.9 release
  \item Authentication using external service (i.e. LDAP)
  \item An option for no authentication
\end{itemize}

\subsection{P1 Requirements}
\begin{itemize}
  \item Custom Java Class authentication options
\end{itemize}


\section{Design}

\subsection{Authentication}

Authentication is simply resolving whether a user and password combinations
corresponds to a valid user.  However, there are several options for validating
this correct username, password combination.  There are currently 4 types of
authentication mechanism in Splice Machine.
\begin{enumerate}
        \item None: All usernames and password are allowed to connect to database.
        \item Native: Usernames in a database table are validated against their
        encrypted password.
        \item LDAP: Usernames checked against an existing LDAP service.
        \item Custom: Custom Java Class provided to perform authentication against the
        database.
\end{enumerate}

\subsubsection{None}
An open database that will allow each user to authenticate against the database
can be configured by setting the following configuration.

\begin{enumerate}
        \item splice.authentication=NONE
\end{enumerate}

This is not recommended for production use.


\subsubsection{Native (default)}
This approach to authentication utilizes the sys.sysusers table in the splice
schema.  The password is encrypted using MD5, SHA-256, or
SHA-512 (Default) algorithm.   

\begin{enumerate}
        \item splice.authentication=NATIVE
      \item splice.authentication.native.algorithm=MD5,SHA-256,SHA-512(Default)
\end{enumerate}

Default user:  \emph{splice}    
Default password: \emph{admin}

Using Native mode, you log in using a modified connection string:

\texttt{splice> connect 'jdbc:derby://localhost:1527/splicedb;user=\emph{user};password=\emph{password}'}

Therefore the administrative login is as follows:

\texttt{splice> connect 'jdbc:derby://localhost:1527/splicedb;user=splice;password=admin'}

\subsubsection{LDAP}
This approach to authentication uses an external LDAP server. To enable this mode, change the property value of splice.authentication to LDAP, and set the following properties to appropriate values according to the LDAP server configuration. 
\begin{enumerate}
        \item splice.authentication=LDAP
        \item splice.authentication.ldap.server=localhost:389
        \item splice.authentication.ldap.searchAuthDN
        \item splice.authentication.ldap.searchAuthPW
        \item splice.authentication.ldap.searchBase
        \item splice.authentication.ldap.searchFilter
\end{enumerate}
An embedded LDAP server will normally run in maven after LDAP mode is enabled, for testing purposes. But for now in a short period time, we encountered problems in maven to be compatible with the embedded LDAP plugin. As a temporary workaround, we use an external LDAP server hosted on an AWS instance, at 54.164.38.153:389. 

\subsubsection{Custom}

The custom authentication
configuration allows for any
class provided by the customer
that implements org.apache.derby.authentication.UserAuthenticator
to be applied as long as the following configuration is set.

\begin{enumerate}
        \item splice.authentication=CUSTOM
        \item   splice.authentication.custom.provider=com.splicemachine.derby.authentication.SpliceUserAuthentication (Default)
\end{enumerate}

The interface is extremely simple and requires a developer to implement the
single method:

boolean authenticateUser(String userName,String userPassword,String
databaseName,Properties info) throws SQLException;

\subsubsection{Splice Scripts}

We currently ship sqlshell.sh (two versions - one for standalone, the other for clustered). Now with NATIVE authentication 
turned on, we hard-code a pair of default user and password in these scripts to automatically log in as the administrator.
In the future we could decide whether to leave one script with no hard-coded user and password such that users can manually 
set them. A potential proposal is to have two scripts:  

\begin{itemize}
\item sqlshelladmin.sh - hardcodes the user to splice/admin for administrator access
\item sqlshell.sh - does NOT auto connect.  User manually connects, providing user and password
\end{itemize}

We are open to have other proposals. 

\subsection{Software Packaging for Standalone/Cluster Installation}
\subsubsection{Standalone Installation}
Since NATIVE authentication mode is turned on by default, package should already include splice-site.xml with properties set to NATIVE authentication mode as shown below:
\begin{lstlisting}
  <property> 
    <name>splice.authentication</name> 
    <value>NATIVE</value> 
  </property> 
  <property> 
    <name>splice.authentication.native.algorithm</name> 
    <value>SHA-512</value> 
  </property> 
\end{lstlisting}

\subsubsection{Cluster Installation}
Package should NOT include splice-site.xml file.

\subsection{Standalone/Cluster Configuration}
Users can toggle between authentication modes by modifying configurations and restarting HBase after configrations are updated. Detail configuration settings for using NONE, NATIVE, and LDAP authentication are listed below. 

\subsubsection{Standalone Configuration}
For NONE authentication mode, edit \\ structured\_derby/src/main/resources/splice-site.xml \\ 
such that it contains the following:
\begin{lstlisting}
  <property> 
    <name>splice.authentication</name> 
    <value>NONE</value> 
  </property> 
\end{lstlisting}
For NATIVE authentication mode, edit \\ structured\_derby/src/main/resources/splice-site.xml \\ 
such that it contains the following: 
\begin{lstlisting}
  <property> 
    <name>splice.authentication</name> 
    <value>NATIVE</value> 
  </property> 
  <property> 
    <name>splice.authentication.native.algorithm</name> 
    <value>SHA-512</value> 
  </property> 
\end{lstlisting}
For LDAP authentication mode, edit \\ structured\_derby/src/main/resources/splice-site.xml \\ such that it contains the following: 
\begin{lstlisting}
  <property> 
    <name>splice.authentication</name> 
    <value>LDAP</value> 
  </property> 
  <property> 
    <name>splice.authentication.ldap.server</name> 
    <value>ldap://54.164.38.153:389/</value> 
  </property> 
  <property> 
    <name>splice.authentication.ldap.searchBase</name> 
    <value>ou=People,dc=splicemachine,dc=com</value> 
  </property> 
\end{lstlisting}

\subsubsection{Cluster Configuration}
If Cloudera Manager is used, add the following properties to hbase-$>$configuration-$>$Service-Wide-$>$Advanced-$>$HBase Service Configuration Safety Valve for hbase-site.xml. Otherwise add the properties to hbase-site.xml manually. \\ \\
For NONE authentication mode:
\begin{lstlisting}
  <property> 
    <name>splice.authentication</name> 
    <value>NONE</value> 
  </property> 
\end{lstlisting}
For NATIVE authentication mode: 
\begin{lstlisting}
  <property> 
    <name>splice.authentication</name> 
    <value>NATIVE</value> 
  </property> 
  <property> 
    <name>splice.authentication.native.algorithm</name> 
    <value>SHA-512</value> 
  </property> 
\end{lstlisting} 
For LDAP authentication mode: 
\begin{lstlisting}
  <property> 
    <name>splice.authentication</name> 
    <value>LDAP</value> 
  </property> 
  <property> 
    <name>splice.authentication.ldap.server</name> 
    <value>ldap://54.164.38.153:389/</value> 
  </property> 
  <property> 
    <name>splice.authentication.ldap.searchBase</name> 
    <value>ou=People,dc=splicemachine,dc=com</value> 
  </property> 
\end{lstlisting}
 

\subsection{Authorization}
Splice Machine will utilize derby authorization syntax and levels. 

\subsubsection{Managing Users}
Users are managed for now using the standard derby syscs calls, for example:

\texttt{splice> call syscs\_util.syscs\_create\_user('username','password');}

\subsubsection{Grant and Revoke}

Splice Machine will utilize derby grant and revoke commands for
authorization levels.  These can be described in the following page.

\url{https://db.apache.org/derby/docs/10.9/devguide/cdevcsecuregrantrevokeaccess.html}

\subsubsection{Roles}

Splice Machine will utilize derby's role based model.  This model is described
in the following page.

\url{https://db.apache.org/derby/docs/10.9/devguide/cdevcsecureroles.html}


\section{Assumptions}

None

\section{Outstanding Issues}

See Upgrade Impacts.  Additionally - what happens if someone wants to "turn off" or "turn on" Authentication after starting the other way around?

\section{Impacts}

\subsection{Impact on QA}

Existing tests need to be updated to reflect an appropriate connection mechanism (splice/admin) or set authentication to NONE

\subsection{Impact on Documentation} 

Existing docs need to be updated if there are references to use of the connection string. In particular, with standalone setting, we have made changes 
to splice-site.xml with several authentication settings. 

\subsection{Impact on Upgrades}

An existing database should be easily migratable to a database enforcing authentication (how?)

\subsection{Other Known Impacts}

See script updates in design above.  Additionally packaging will need to be updated to set the default settings:
\begin{itemize}
\item For standalone, this will go into splice-site.xml.
\item For clustered, we will document how to pass these parameters in with other splice-specific parameters that go into hbase-site.xml
\end{itemize}
 

\end{document}

